### 2020-09-01 response.tex

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{xspace}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}

\newcommand{\rR}{\mbox{$r$--$\cal R$}}
\newcommand{\RR}{\ensuremath{{\cal R}}}
\newcommand{\RRhat}{\ensuremath{{\hat \cal R}}}
\newcommand{\Rx}[1]{\ensuremath{{\cal R}_{#1}}} 
\newcommand{\Ro}{\ensuremath{{\mathcal R}_{0}}\xspace}
\newcommand{\Rpool}{\ensuremath{{\mathcal R}_{\textrm{\tiny{pool}}}}\xspace}
\newcommand{\Reff}{\Rx{\mathit{eff}}}
\newcommand{\Tc}{\ensuremath{C}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript. 
We have made major revisions to our manuscript to address the reviewers' comments.
We were apparently not clear in our writing. Reviewers 1 and 3 both pointed out that we assume that the latent and incubation periods are equivalent (and thus not allowing for presymptomatic transmission). 
However, we only present this result in the appendix as a special case.
We do not assume this in the main text and allow for presymptomatic transmission.
Below please find our responses to reviewers.

\rev{Reviewer \#1}

\revtext{I have one main concern, and this concerns the model. First a very general model is discussed, then a very special version is mentioned (mid p11) in which the latent period and incubation periods are always identical. For this model the General interval and Serial interval have exactly the same distribution. My concern is whether the stated results are valid only for this special case or if they apply also in general. I doubt the results hold in general, but at least think arguments for it are missing.}

The special version is only presented in the supplementary material. We do not assume this in the main text. We have made this distinction clear. Also Figure 2C--D shows that the initial forward serial-interval distribution is different from the intrinsic generation-interval distribution.

\revtext{Regarding my main concern above I note that you write "should" on p3, line 91, and "expect" on p 10, line 273. I don't see any proof for the statement to hold in general and am not sure the statement is even true in general. }

We provide a mathematical proof of this using a general model in the supplementary file. Figure 2A also demonstrates that this holds. We tried to make this clear in the main text.

\revtext{p4, bottom: refer to Fig 1 }

\revtext{p5, middle: emphasize that s and p are calendar times whereas $\tau$ is a duration }

\revtext{p5, line 158: clarify that $\tau = s-p$}

Done.

``Consequently, no matter how delays are distributed, if
$\mathcal P$ and $\mathcal S$ represent the sizes of primary and
secondary cohorts then we can express the total density of intervals $\tau$ between calendar time $\psymp$ and $\ssymp$ (i.e., $\tau=\ssymp-\psymp$) as follows:
\begin{equation}
W(\psymp) \psize(\psymp) f_\psymp(\tau) = \ssize(\ssymp) b_\ssymp(\tau) \,,
\label{eq:match}
\end{equation}''

\revtext{p5, line 170: should be $W(s-\tau)$?}

Done.

\revtext{p6, line 181-183: why will it contract during the outbreak? I understand why fewer will be materialized, but not why they will get shorter}

We have tried to clarify this:

``For example, shorter generation intervals are more likely to be realized later in the \emph{epidemic} (therefore, leading to ``contraction'' of generation intervals) because infected individuals are less likely to infect others later in their \emph{infection} due to factors that drive time-dependent decreases in transmission such as susceptible depletion, behavioral change, and interventions \citep{champredon2015intrinsic}.''

We also demonstrate this phenomenon in Figure 3C.

\revtext{p6, line 188: shorter than what? If you mean compared to the forward delay distribution it is always true when r>0 and not necessary that r is higher (than what?).}

We have rephrased this sentence:

``Thus, as the growth rate $r$ of an epidemic increases, the mean backward delay distribution will decrease, leading to stronger bias.''

\revtext{p7, line 203-205. Several of these papers acknowledge that the means are identical only if the tau's have the same distribution. On p11 (cf my main concern) you study an even more restrictive situation so in a sense you also implicitly assume this to hold or have I misunderstood?}

The restrictive is presented in the supplementary file as a special case. The main results do not depend on the restrictive assumption. We have tried to clarify this. We have also re-phrased the sentence now to acknoweldge what other studies have said more carefully:

``These studies concluded that the serial and generation intervals have the same mean when $\tau_{\rm i1}$ and $\tau_{\rm i2}$ are drawn from the same distributions \citep{svensson2007note,klinkenberg2011correlation,champredon2018equivalence, britton2019estimation};
however, distributions of realized incubation periods, $\tau_{\rm i1}$ and $\tau_{\rm i2}$ will be identical only if we assume that they are intrinsic to individuals (and not dependent on epidemic dynamics at the population-level) ---
something that is generally true of forward but not backward incubation-period distributions.''

\revtext{Section 3: be specific which model you are assuming. I think it is the SEIR where incubation is identical to latent period, i.e. individuals become infectious the same instance as their onset of symptoms.}

We are not assuming this. We have tried to clarify this in the main text.

\revtext{p19, line 470: something missing}

We have re-phrased the sentence to make it clearer:

``For example, \cite{thompson2019improved} recently emphasized the importance of using up-to-date serial-interval data for accurate estimation of time-varying reproduction numbers.
However, our results show that if changes the forward serial interval through time are not accounted for, using up-to-date serial-interval data can, in fact, exacerbate the underestimation of initial \RR.''

\revtext{p19, line 485: you are assuming generation and serial intervals have the same distribution.}

We do not assume this. Also, Figure 2--3 illustrate that generation and serial intervals have different means, and therefore different distributions. We have tried to highlight these differences in the main text.

\rev{Reviewer \#2}

\revtext{The authors show how methods for estimating the reproduction number R from the growth rate r based on incorrectly defined disease transmission intervals can lead to biased estimates. They explain where the biases come from and present an unbiased heuristic to derive R from r, which they apply to the COVID-19 outbreak in China. Because the reproduction number has become part of everyday vocabulary this spring, the explanation of the source of biases is sound and clear, and the presented heuristic likely useful, their analysis may be sufficiently important to warrant publication in PNAS. However, then the authors must substantiate their criticism of previous work and present the reproduction number in a nuanced context.}

\revtext{The methods and results sections do excellent jobs explaining the source of confusion and biases, and how to overcome them. The discussion section synthesizes the insights into suggestions and good practices. However, I miss examples of how far off - if off at all - criticized application papers are when estimating the reproduction number.}

\revtext{The authors can improve their manuscript by responding to the following issues:}

\revtext{What is the manuscript about - generation and serial intervals or the reproduction number R? The title and abstract suggest the intervals and the introduction the reproduction number. I think opening with the well-known reproduction number will give the most transparent and coherent story.}

It is about both. Generation and serial intervals are key quantities in describing epidemic dynamics; their differences have long been misunderstood. These intervals also play critical role in estimating the reproduction number. We have made major revisions to the writing of the manuscript to provide a transparent and coherent story.

\revtext{The reproduction number is not put in a nuanced context. The presentation of the reproduction number as "one of the most important characteristics of an emerging epidemic ... [definition] ... [that] allows us to predict the extent to which an infection will spread in the population, and the amount of intervention necessary to eliminate it" that then leads directly into how it is estimated leaves no room for its shortcomings. A more nuanced picture includes its inability to, for example, account for individual variation, including superspreaders and spatial variations in regions.}

\revtext{The introduction presents previous methodological work on the R-r relationship based on generation and serial intervals and recent applications that confuse the two intervals. The presentation goes back and forth between the two and does not clarify the shortcomings of each work or application. My suggestion for a stronger introduction: A funnel focusing on the methods that ends with the knowledge gap "In contexts where the distributions are expected to be different, current theory has no explanation for how these differing distributions could provide identical estimates of R from r" (and further specified by the gaps indicated at the end of section 2.1 and beginning of section 2.2) followed by concrete examples of how off R estimates are in applications to substantiate the problem.}

\revtext{I also lack a discussion of standard software to estimate the reproduction number such as the widely used R package EpiEstim.}

\revtext{The abstract states that "recent studies suggest that the two intervals give different estimates of R from r" and "our analysis shows that using incorrectly defined serial intervals can severely bias estimates" Exemplify by numbers to substantiate the severity.}

\revtext{Same vagueness at the end of the introduction: "Conversely, using inaccurately defined serial intervals or failing to account for changes in the observed serial-interval distributions over the course of an epidemic can considerably bias estimates of R." By how much?}

\revtext{Figure 1 clearly illustrates the different intervals from different perspectives but does not explain the source of biases. Consider ways to visually communicate the important insights at the end of section 3.1.}

\revtext{Section 3.3 provides one example of a biased R estimate. The authors should substantiate their critique of several papers in the introduction by listing their estimated R estimates and the revised estimates - or not accusing those papers in the introduction. I think "these studies should also re-assess whether they appropriately considered the forward serial interval" is weak.}

\rev{Reviewr \#3}

\revtext{Throughout the paper becoming symptomatic seems equivalent to becoming infectious? This is a little confusing in context of Covid-19.}

We do not assume this. This is only assumed in the supplementary file as a special case. We now make this clear in the main text.

\revtext{Define s=p+tau before eq. 1.}

Done.

``Consequently, no matter how delays are distributed, if
$\mathcal P$ and $\mathcal S$ represent the sizes of primary and
secondary cohorts then we can express the total density of intervals $\tau$ between calendar time $\psymp$ and $\ssymp$ (i.e., $\tau=\ssymp-\psymp$) as follows:
\begin{equation}
W(\psymp) \psize(\psymp) f_\psymp(\tau) = \ssize(\ssymp) b_\ssymp(\tau) \,,
\label{eq:match}
\end{equation}''

\revtext{On page 9, ``any deterministic epidemic model"...Guess it should be just as important for a stochastic model?}

We removed the word ``deterministic'' from the sentence.

\revtext{On page 11, the important parameter rho =-0.5, 0 or 0.5 is obscure. What does these correlations means? partly anti-correlated, non correlated or positive correlated, but by how much? Which disease could have anti-correlation between these quantities? It make interpretation of Fig. 2B difficult. }

\revtext{Page 4, figure 3A: Why are stochastic trajectories systematically shifted toward earlier time compared to deterministic trajectory. Was the deterministic trajectory not simulated with same serial interval distribution?}

\revtext{Page 18, ``based on 8 and 6 doubling periods" guess it should be "based on a doubling period of 8 or 6 days"??}

Thank you. We have changed this as suggested.

\revtext{Page 17, figure 5 is difficult to judge. It needs to be supplemented with a trajectory of the epidemic curve in the days that is shown.}

\revtext{Figure 5, what is on the y axis? Which delay is plotted and how does it relate to the forward/backward serial intervals. Why not plot serial intervals directly?}

\revtext{Could the results be simplified to an extend that medical doctors that actually "measure" backward serial intervals directly could recalculate the intrinsic serial intervals (in some simple approximation). As a technical paper it should provide some direct applicable results.}

\revtext{Overall, I find the work interesting and believe it could contribute to current discussions on intrinsic time scales for Covid-19. It is a technical read, that would gain by being made easier accessible.}

Thank you. We have made substantial edits to the manuscript to make it more accessible.

\bibliography{serial}

\end{document}

### 2020-09-15 serial.tex
update

To calculate realized serial-interval distributions, we begin by modeling $\total(\psymp,\ssymp)$: the total density of serial intervals that start (when infectors develop symptoms) at time $\psymp$ and end (when infectees develop symptoms) at time $\ssymp$.
The density of serial intervals between time $\psymp$ and $\ssymp$,
given that the infectors became infected at time $\pinf\le\psymp$ and
the infectees became infected at time $\sinf\le\ssymp$, depends on the
amount of infection that occurs between time $\pinf$ and $\sinf$ as
well as the density of forward incubation periods between $\pinf$ and
$\psymp$ (realized incubation periods of infectors) and between
$\sinf$ and $\ssymp$ (realized incubation periods of infectees):
\begin{equation}
\underbrace{\Rc (\pinf)}_{\substack{\text{case} \\ \text{reproduction} \\ \text{number}}} 
\times 
\underbrace{i(\pinf)}_{\text{incidence}} 
\times 
\underbrace{h_{\pinf}(\psymp-\pinf, \sinf - \pinf)}_{\substack{\text{joint density of} \\ \text{the forward incubation} \\ \text{period } p-\pinf \text{ and the forward} \\ \text{generation interval } \sinf-\pinf\\ \text{(of infectors)}}}
\times
\underbrace{\idist_{\sinf}(\ssymp - \sinf)}_{\substack{\text{marginal density of} \\ \text{the forward incubation} \\ \text{period } \ssymp-\sinf \\ \text{(of infectees)}}},
\end{equation}
where the case reproduction number $\Rc (\pinf)$ is defined as the average number of secondary cases that a primary case infected at time $\pinf$ will generate over the course of their infection \citep{fraser2007estimating}.
We describe the forward incubation periods and the forward generation intervals using a joint probability distribution because onset of symptoms and transmission potential jointly depend on the life history of a disease;
for example, if an infected individual can only transmit the disease after symptom onset, the forward generation interval will necessarily be longer than the forward incubation period.

@@

We see that the denominator of the forward serial-interval distribution, $\int_{-\infty}^\infty \total(\psymp, \psymp+\tau') \dtau'$, then corresponds to the total number of symptomatic cases generated by primary cases who developed at time \psymp (\eref{fserial});
dividing it by the incidence of symptomatic cases (also infected by symptomatic cases) at time \psymp, $j(\psymp) = \int_{-\infty}^\infty \total(\psymp-\tau', \psymp) \dtau'$, we obtain the serial reproduction number:
\begin{equation}
\Rs(\psymp) = \frac{\int_{-\infty}^\infty \total(\psymp, \psymp+\tau') \dtau'}{j(\psymp)},
\end{equation}
which we define as the average number of symptomatic cases generated by individuals who developed symptoms at time \psymp (and were also infected by sympmtomatic cases).
Combining the forward serial-interval distribution with the serial reproduction number completes the renewal process between symptomatic cases:
\begin{equation}
j(t) = \int_{-\infty}^\infty \Rs(t-\tau) j(t-\tau) f_{t-\tau}(\tau) \dtau.
\label{eq:jrenew}
\end{equation}
This framework allows us to understand changes in the realized serial intervals for any epidemic model and properly link serial-interval distributions with the renewal process.

@@

We illustrate changes in the forward and backward serial intervals over the course of an epidemic by applying it to a specific example of an epidemic model.
We model disease spread with a renewal-equation model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
Ignoring births and deaths, general changes in the proportion of susceptible individuals $S(t)$ and incidence of infection $i(t)$ can be described as:
\begin{align}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - i(t)\nonumber\\
i(t) &= \RR(t) \int_0^\infty i(t-\tau) \gdist(\tau) \dtau,
\label{eq:renewal}
\end{align}
where $\RR(t)$ is the instantaneous reproduction number (i.e., the average number of secondary cases that a primary case infected at time $t$ will generate if conditions at time $t$ remains unchanged \citep{fraser2007estimating}), and $\gdist(\tau)$ is the intrinsic generation-interval distribution (i.e., the forward generation-interval distribution of a primary case in a population where changes in $\RR(t)$ is negligible \citep{champredon2015intrinsic}).
This model effectively assumes that epidemic dynamics are driven by changes in transmission rate, which may not hold under individual-based intervention such as case isolation \citep{fraser2007estimating};
nonetheless, this form has been widely used in the literature and has been successfully applied in modeling the current COVID-19 pandemic.

Here, we consider a specific example where changes in the reproduction number is driven by susceptible depletion: $\RR(t) = \Ro S(t)$, where \Ro represents the basic reproduction number.
Then, the forward generation-interval for a cohort of individuals that were infected at time $\psymp$ follows \citep{champredon2015intrinsic}:
\begin{equation}
\gdist_\psymp (\tau) = \frac{ \gdist(\tau) S(\psymp + \tau)}{\int_0^\infty \gdist(\tau') S(\psymp + \tau') \dtau'},
\end{equation}
which allows us to separate the joint probability distribution $h_\psymp$ of the forward incubation period and the forward generation-interval distribution as a product of the proportion of susceptible individuals $S$ and the joint probability distribution $h$ of the forward incubation period and the intrinsic generation intervals:
\begin{equation}
h_\psymp (x, \tau) = \frac{h(x, \tau) S(\psymp + \tau)}{\int_0^\infty \int_0^\infty h(x', \tau') S(\psymp + \tau') \dtau' \dx'}.
\end{equation}
We further assume that the forward incubation period distribution does
not vary across cohorts over the course of an epidemic, as it
represents the life history of a disease; we denote it as $\idist$.
Then, we have:
\begin{align}
\idist(x) &= \int_0^\infty h(x, \tau) \dtau\nonumber,\\
\gdist(\tau) &= \int_0^\infty h(x, \tau) \dx.
\label{eq:marginal}
\end{align}
Finally, the case reproduction for this model is defined as follows:
\begin{equation}
\Rc(t) = \Ro \int_0^\infty \gdist(\tau) S(t+\tau) \dtau.
\end{equation}
The forward and backward serial-interval distributions are then calculated by substituting these quantities into \eref{fserial} and \eref{bserial}.
We use this framework to illustrate how the realized epidemiological time distributions vary over the course of an epidemic and depend on the perspective (i.e., forward vs. backward).

@@

During the initial phase of an epidemic, the proportion susceptible remains approximately constant ($S(t) \approx S(0)$) and incidence of infection grows exponentially: $i(t) \approx i_0\exp(rt)$.
During this period, the intrinsic generation-interval distribution provides the correct link between the exponential growth rate $r$ and the initial reproduction number $\RR=\Ro S(0)$ based on the Euler-Lotka equation \citep{wallinga2007generation}.
Here, we focus on the estimates of the basic reproduction number \Ro (the value of \RR in a fully susceptible population, $S(t) \approx 1$):
\begin{equation}
\frac{1}{\Ro} = \int_0^\infty \exp(-r\tau) g(\tau) \dtau.
\label{eq:Rgen}
\end{equation}
Analogous to the intrinsic generation-interval distribution, 
forward serial-interval distributions describe the renewal process of infection based on symptom onset dates.
Therefore, we expect the forward serial-interval distribution during the exponential growth phase --- which we refer to as the \emph{initial} forward serial-interval distribution $f_0$ --- to estimate the same value of \Ro for a given $r$ as the intrinsic generation-interval distribution (note, however, that the forward serial interval is not necessarily positive):
\begin{equation}
\frac{1}{\Ro} = \int_{-\infty}^\infty \exp(-r\tau) f_{0}(\tau) \mathrm{d} \tau,
\label{eq:Rforward}
\end{equation}
where the initial forward serial-interval distribution is given by:
\begin{equation}
f_{0}(\tau) = \frac{1}{\phi} \int_{-\infty}^{0} \int_{\pinf}^{\tau} \exp(r \pinf) h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf,
\label{eq:initialSI}
\end{equation}
where the normalization constant $\phi$ is determined by the
requirement that $\int_{-\infty}^\infty f_{0}(\tau)\,\dtau=1$.
Since we have not yet made any assumptions about the shape of the joint distribution $h$ between incubation periods and the generation intervals, this conclusion is robust, regardless of whether presymptomatic is possible or not.
It is straightforward to see this relationship from \eref{jrenew} (by substituting $j(t) \approx j(0) \exp(rt)$ and $\Rs(t) \approx \Ro$);
in Supplementary Materials, we provide a more detailed mathematical proof of this relationship.
We further compare this with the estimate of \Ro based on the intrinsic serial-interval distribution $q(\tau)$:
\begin{equation}
\frac{1}{\Rintrinsic} = \int_{-\infty}^\infty \exp(-r\tau) q(\tau) \mathrm{d} \tau.
\label{eq:Rintrinsic}
\end{equation}
Here, the intrinsic serial-interval distribution $q(\tau)$ does not depend on epidemic dynamics:
\begin{equation}
q(\tau) = \frac{1}{\phi_q} \int_{-\infty}^{0} \int_{\pinf}^{\tau} h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf,
\label{eq:intrinsicSI}
\end{equation}
where the normalization constant $\phi_q$ is determined by the
requirement that $\int_{-\infty}^\infty q(\tau)\,\dtau=1$.
Rather than numerically integrating over closed forms of $g$, $f_0$, and $q$ to estimate $\Ro$, we use simulation-based approaches for simplicity (Supplementary Materials).

@@

The initial forward serial-interval distribution captures the exponential growth phase of an epidemic.
We now explore how forward and backward serial intervals can vary over the course of an epidemic using deterministic and stochastic simulations based on the renewal equations (see Supplementary Materials) using parameters in Table 1;
we further assume $\Ro=2.5$ to reflect the transmission dynamics of COVID-19 \citep{doi:10.1098/rsif.2020.0144}.
While the forward serial-interval distribution is our primary focus, understanding the differences between the forward and the backward distributions is important because the observed intervals during an ongoing epidemic are often the backward ones:
we typically identify infected individuals and ask when and by whom they were infected.
Similarly, when we are estimating the incubation period of an individual, we typically observe their symptom onset date and try to estimate when they were infected (e.g., \cite{backer2020incubation}).

@@

@@

Now, we turn to issues of estimating the reproduction number from the observed serial-interval data during on ongoing epidemic.
In order to have an unbiased estimate of the basic reproduction number, we need to estimate the initial forward serial-interval distribution --- i.e., estimate based on cohorts of infectors who share the same symptom onset time, at the early stage of the epidemic.
However, researchers typically use all available information to estimate epidemiological parameters;
in particular, \cite{thompson2019improved} recently suggested that up-to-date serial-interval data are necessary to accurately estimate the reproduction number.
We explore the consequences of neglecting changes in the realized serial-interval distribution on estimates of the basic reproduction number.

>>>>

To calculate realized serial-interval distributions, we begin by modeling $\total(\psymp,\ssymp)$: the total density of serial intervals that start (when infectors develop symptoms) at time $\psymp$ and end (when infectees develop symptoms) at time $\ssymp$.
For simplicity, we assume that all infected individuals eventually develop symptoms.
Then, the density of serial intervals between time $\psymp$ and $\ssymp$,
given that the infectors became infected at time $\pinf\le\psymp$ and
the infectees became infected at time $\sinf\le\ssymp$, depends on the
amount of infection that occurs between time $\pinf$ and $\sinf$ as well as
the density of forward incubation periods between $\pinf$ and
$\psymp$ (realized incubation periods of infectors) and between
$\sinf$ and $\ssymp$ (realized incubation periods of infectees):
\begin{equation}
\underbrace{\Rc (\pinf)}_{\substack{\text{case} \\ \text{reproduction} \\ \text{number}}} 
\times 
\underbrace{i(\pinf)}_{\substack{\text{incidence} \\ \text{of} \\ \text{infection}}} 
\times 
\underbrace{h_{\pinf}(\psymp-\pinf, \sinf - \pinf)}_{\substack{\text{joint density of} \\ \text{forward incubation} \\ \text{periods } p-\pinf \text{ and forward} \\ \text{generation intervals } \sinf-\pinf\\ \text{(of infectors)}}}
\times
\underbrace{\idist_{\sinf}(\ssymp - \sinf)}_{\substack{\text{marginal density of} \\ \text{forward incubation} \\ \text{periods } \ssymp-\sinf \\ \text{(of infectees)}}},
\end{equation}
where the case reproduction number $\Rc (\pinf)$ is defined as the average number of secondary cases that a primary case infected at time $\pinf$ will generate over the course of their infection \citep{fraser2007estimating}.
We describe the forward incubation periods and the forward generation intervals using a joint probability distribution because onset of symptoms and transmission potential jointly depend on the life history of a disease;
for example, if an infected individual can only transmit the disease after symptom onset, the forward generation interval will necessarily be longer than the forward incubation period.

@@

We see that the denominator of the forward serial-interval distribution, $\int_{-\infty}^\infty \total(\psymp, \psymp+\tau') \dtau'$, then corresponds to the total number of symptomatic cases generated by primary cases who themselves developed symptoms at time \psymp\ (\eref{fserial}).
Dividing this quantity by the number of individuals who developed symptoms at time at time \psymp, $j(\psymp) = \int_{-\infty}^\infty \total(\psymp-\tau', \psymp) \dtau'$, we obtain the serial reproduction number:
\begin{equation}
\Rs(\psymp) = \frac{\int_{-\infty}^\infty \total(\psymp, \psymp+\tau') \dtau'}{j(\psymp)},
\end{equation}
which we define as the average number of symptomatic cases generated by individuals who developed symptoms at time \psymp.
Combining the forward serial-interval distribution with the serial reproduction number completes the renewal process between symptomatic cases:
\begin{equation}
j(t) = \int_{-\infty}^\infty \Rs(t-\tau) j(t-\tau) f_{t-\tau}(\tau) \dtau.
\label{eq:jrenew}
\end{equation}
This framework allows us to understand changes in the realized serial intervals for any epidemic model and properly link serial-interval distributions with the renewal process.

@@

We illustrate changes in forward and backward serial intervals over the course of an epidemic by applying it to a specific example of an epidemic model.
We model disease spread with a renewal-equation model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
Ignoring births and deaths, general changes in the proportion of susceptible individuals $S(t)$ and incidence of infection $i(t)$ can be described as:
\begin{align}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - i(t)\nonumber\\
i(t) &= \RR(t) \int_0^\infty i(t-\tau) \gdist(\tau) \dtau,
\label{eq:renewal}
\end{align}
where $\RR(t)$ is the instantaneous reproduction number (i.e., the average number of secondary cases that a primary case infected at time $t$ will generate if conditions at time $t$ remains unchanged \citep{fraser2007estimating}), and $\gdist(\tau)$ is the intrinsic generation-interval distribution (i.e., the forward generation-interval distribution of a primary case in a population where changes in $\RR(t)$ is negligible \citep{champredon2015intrinsic}).
This model effectively assumes that epidemic dynamics are driven by changes in transmission rate, which may not hold under individual-based intervention such as case isolation \citep{fraser2007estimating};
nonetheless, this form has been widely used in the literature and has been successfully applied in modeling the current COVID-19 pandemic \citep{gostic2020practical}.

Here, we consider a specific example where changes in the reproduction number is driven by susceptible depletion: $\RR(t) = \Ro S(t)$, where \Ro represents the basic reproduction number.
Although we do not consider non-pharmaceutical interventions explicitly, the effect of susceptible depletion is mathematically equivalent to the effect of interventions that reduce transmission rate.
Then, the forward generation-interval for a cohort of individuals that were infected at time $\psymp$ follows \citep{champredon2015intrinsic}:
\begin{equation}
\gdist_\psymp (\tau) = \frac{ \gdist(\tau) S(\psymp + \tau)}{\int_0^\infty \gdist(\tau') S(\psymp + \tau') \dtau'},
\end{equation}
which allows us to separate the joint probability distribution $h_\psymp$ of the forward incubation period and the forward generation-interval distribution as a product of the proportion of susceptible individuals $S$ and the joint probability distribution $h$ of the forward incubation period and the intrinsic generation intervals:
\begin{equation}
h_\psymp (x, \tau) = \frac{h(x, \tau) S(\psymp + \tau)}{\int_0^\infty \int_0^\infty h(x', \tau') S(\psymp + \tau') \dtau' \dx'}.
\end{equation}
We further assume that the forward incubation period distribution does
not vary across cohorts over the course of an epidemic, as it
represents the life history of a disease; we denote it as $\idist$.
Then, we have:
\begin{align}
\idist(x) &= \int_0^\infty h(x, \tau) \dtau\nonumber,\\
\gdist(\tau) &= \int_0^\infty h(x, \tau) \dx.
\label{eq:marginal}
\end{align}
Finally, the case reproduction for this model is defined as follows:
\begin{equation}
\Rc(t) = \Ro \int_0^\infty \gdist(\tau) S(t+\tau) \dtau.
\end{equation}
The forward and backward serial-interval distributions are then calculated by substituting these quantities into \eref{fserial} and \eref{bserial}.
We use this framework to illustrate how the realized epidemiological time distributions vary over the course of an epidemic and depend on the perspective (i.e., forward vs. backward).

@@

During the initial phase of an epidemic, the proportion susceptible remains approximately constant ($S(t) \approx S(0)$) and incidence of infection grows exponentially: $i(t) \approx i_0\exp(rt)$.
During this period, the intrinsic generation-interval distribution provides the correct link between the exponential growth rate $r$ and the initial reproduction number $\RR=\Ro S(0)$ based on the Euler-Lotka equation \citep{wallinga2007generation}.
Here, we focus on the estimates of the basic reproduction number \Ro (the value of \RR in a fully susceptible population, $S(t) \approx 1$):
\begin{equation}
\frac{1}{\Ro} = \int_0^\infty \exp(-r\tau) g(\tau) \dtau.
\label{eq:Rgen}
\end{equation}
Analogous to the intrinsic generation-interval distribution, 
forward serial-interval distributions describe the renewal process of infection based on symptom onset dates.
Therefore, we expect the forward serial-interval distribution during the exponential growth phase --- which we refer to as the \emph{initial} forward serial-interval distribution $f_0$ --- to estimate the same value of \Ro for a given $r$ as the intrinsic generation-interval distribution (note, however, that the forward serial interval is not necessarily positive):
\begin{equation}
\frac{1}{\Ro} = \int_{-\infty}^\infty \exp(-r\tau) f_{0}(\tau) \mathrm{d} \tau,
\label{eq:Rforward}
\end{equation}
where the initial forward serial-interval distribution is given by:
\begin{equation}
f_{0}(\tau) = \frac{1}{\phi} \int_{-\infty}^{0} \int_{\pinf}^{\tau} \exp(r \pinf) h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf,
\label{eq:initialSI}
\end{equation}
where the normalization constant $\phi$ is determined by the
requirement that $\int_{-\infty}^\infty f_{0}(\tau)\,\dtau=1$.
Since we have not yet made any assumptions about the shape of the joint distribution $h$ between incubation periods and the generation intervals, this conclusion is robust, regardless of whether presymptomatic is possible or not.
It is straightforward to see this relationship from \eref{jrenew} (by substituting $j(t) \approx j(0) \exp(rt)$ and $\Rs(t) \approx \Ro$);
in Supplementary Materials, we provide a more detailed mathematical proof of this relationship.

We further compare this with the estimate of \Ro based on the intrinsic serial-interval distribution $q(\tau)$:
\begin{equation}
\frac{1}{\Rintrinsic} = \int_{-\infty}^\infty \exp(-r\tau) q(\tau) \mathrm{d} \tau.
\label{eq:Rintrinsic}
\end{equation}
Here, the intrinsic serial-interval distribution $q(\tau)$ does not depend on epidemic dynamics:
\begin{equation}
q(\tau) = \frac{1}{\phi_q} \int_{-\infty}^{0} \int_{\pinf}^{\tau} h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf,
\label{eq:intrinsicSI}
\end{equation}
where the normalization constant $\phi_q$ is determined by the
requirement that $\int_{-\infty}^\infty q(\tau)\,\dtau=1$.
Rather than numerically integrating over closed forms of $g$, $f_0$, and $q$ to estimate $\Ro$, we use simulation-based approaches for simplicity (Supplementary Materials).

@@

The initial forward serial-interval distribution captures the exponential growth phase of an epidemic.
We now explore how forward and backward serial intervals can vary over the course of an epidemic using deterministic and stochastic simulations based on the renewal equations (see Supplementary Materials) using parameters in Table 1;
we further assume $\Ro=2.5$ to reflect the transmission dynamics of COVID-19 in China \citep{doi:10.1098/rsif.2020.0144}.
While the forward serial-interval distribution is our primary focus, understanding the differences between the forward and the backward distributions is important because the observed intervals during an ongoing epidemic are often the backward ones:
we typically identify infected individuals and ask when and by whom they were infected.
Similarly, when we are estimating the incubation period of an individual, we typically observe their symptom onset date and try to estimate when they were infected (e.g., \cite{backer2020incubation}).

@@

Stochastic simulations confirm our deterministic calculations albeit minor differences due to transient dynamics.
Incidence trajectories of stochastic simulations are slightly shifted (\fref{epi}A) due to differences in initial conditions. 
We set the distribution of infection time of initially infected individuals (beginning from a small number of infected individuals) for deterministic simulations such that the system can exhibit exponential growth immediately (Supplementary Materials).
This is not possible for stochastic simulations because the number of initially infected individuals cannot be smaller than 1; instead, we begin with 10 infected indiivduals who are infected at time $t=0$.
In addition, the mean backward intervals of stochastic simulations begins from smaller values due to left-censoring.

@@

Now, we turn to practical issues of estimating the reproduction number from the observed serial-interval data during on ongoing epidemic.
In order to have an unbiased estimate of the basic reproduction number, we need to estimate the initial forward serial-interval distribution --- i.e., serial intervals based on cohorts of infectors who share the same symptom onset time, at the early stage of the epidemic.
However, researchers typically use all available information to estimate epidemiological parameters (e.g., aggregating all serial intervals observed until certain time of an epidemic);
for example, \cite{thompson2019improved} recently suggested that up-to-date serial-interval data are necessary to accurately estimate the reproduction number.
We explore the consequences of neglecting changes in the realized serial-interval distribution on estimates of the basic reproduction number.
